Evolution, functional diversity, regime shifts, hysteresis, and nonlinearity; ecosystem responses to environmental change.

### 2021-05-17 manuscript/article.Rmd
article work, mostly

When thinking about ecosystem stability, a key question is how biodiversity influences ecosystem responses to environmental change (figure 1: panel a shows linear response with different gradients, large being with low diversity; panel b shows a linear and a non-linear response starting and ending at the same place; panel c shows a hysteretic response, with greater or less range of hysteresis).

In this context, a directly relevant aspect of biodiversity is variation in functional traits, i.e. the traits that determine organismal responses to environmental conditions and that determine the effects of organisms on their environment (citation). Variation in functional traits is often termed "functional diversity".

There is a considerable amount of theory predicting, and evidence supporting, that greater functional diversity leads to lower temporal variability in ecosystem states variables (such as total biomass) (citations). The mechanism by which this occurs is that greater functional response diversity causes a weaker relationship between an environmental driver gradient and the ecosystem state (figure 1a), due to the functional turnover along the environmental driver gradient that results from variation in organisms' responses to the environmental driver gradient .

@@

These explanations/predictions were based on the verbal/graphical argumentation @dakos2019b. There were no mathematical models of ecosystem responses to environmetnal change behind the explanations/predictions. @ceulemans2021, in their article *Functional diversity alters the effects of a pulse perturbation on the dynamics of tritrophic food webs*, did employ a such a mathematical model.

>>>>

This report concerns how biodiversity influences the strength and nature of ecosystem responses to environmental change (figure 1: panel a shows linear response with different gradients, large being with low diversity; panel b shows a linear and a non-linear response starting and ending at the same place; panel c shows a hysteretic response, with greater or less range of hysteresis). Strong responses of ecosystems to environmental change imply high *sensitivity* and lack of resistance to environmental change, and higher temporal variability in the face of varying environmental conditions. Greater *non-linearity* in ecosystem responses increases the potential for surprises, such as when an ecosystem that was resistant to an environmental change becomes unexpectedly sensitive to further change. The presence and extent of *hysteresis* in ecosystem responses to environmental driver gradients concerns how readily are ecosystem responses reversed when environmental change is reversed.

In this context, a directly relevant aspect of biodiversity is variation in functional traits, i.e. the traits that determine organismal responses to environmental conditions and that determine the effects of organisms on their environment (citation). Variation in functional traits is often termed "functional diversity". There is a considerable amount of theory predicting, and evidence supporting, that greater functional diversity leads to lower temporal variability in ecosystem states variables (such as total biomass) (citations) (Leary, Pennekamp). The mechanism by which this occurs is that greater functional response diversity causes a weaker relationship between an environmental driver gradient and the ecosystem state (figure 1a), due to the functional turnover along the environmental driver gradient that results from variation in organisms' responses to the environmental driver gradient .

@@

These explanations/predictions were based on the verbal/graphical argumentation @dakos2019b. There were no mathematical models of ecosystem responses to environmental change behind the explanations/predictions. @ceulemans2021, in their article *Functional diversity alters the effects of a pulse perturbation on the dynamics of tritrophic food webs*, did employ a such a mathematical model.

We discuss empirical studies in the Discussion section (spoiler--there are few, if any, directly relevant ones).
